### 2016-05-26 syntax.tex
hammering on syntax

\heading{Discusssion.}
This syntax captures classical Bloom filters by setting~$\Sigma=\bits$ and defining $\mathsc{Rep},\mathsc{Qry}$ in the obvious way.  Setting $\Sigma=\bits^{\leq L}$ allows one to store a bitstring of length at most~$L$ at each location. \tsnote{Would let us capture a construction from NY, among other things.}  We note that even though this would allow for the storage of counters (say) at each filter location, our restriction that~$\mathsc{Qry}$ returns a bit does not capture functionality of Bloom filters that employ counters, e.g., returning a counter value.

@@

\heading{Mutuable hash-based filters. }
Fix a nonempty sets $U,\Sigma$ and integers $k,m_1,m_2,>0$ with $m_1 \leq m_2$.  Fix a symbol $\bot \not\in U$.  An $(k,[m_1,m_2])$-filter (over universe~$U$) is a tuple  $B=(\mathsc{H},\mathsc{Rep},\mathsc{Qry},\mathsc{Up}, \mathsc{Test})$.   
%
The randomized \emph{hash-sampling} algorithm~$\mathsc{H}$ samples a size~$k$ subset~$\mathcal{H} \subset \mathrm{Func}(U,\{0,1,\ldots,m_2-1\})$.  We write $\mathcal{H} \getsr \mathsc{H}$ for this operation.
%
The deterministic representation algorithm $\mathsc{Rep}\colon [U]^n \rightarrow \left(\bigcup_{m=m_1}^{m_2}\Sigma^m\right)$ takes a set~$S$ of~$n$ elements as input, and outputs representation~$M$ of length~$m_1 \leq m \leq m_2$.
%\mbox{s.t. } \mu_S(x)>0, \mu_T(x)>0\}$.  On the other hand, we may end up deciding it is overkill... }
%
The determinisitc query algorithm $\mathsc{Qry}\colon \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right)\times U \rightarrow \bits^*$ takes a representation $M$ and an element $x \in U$ and returns a bitstring.  
%
The deterministic \emph{update} algorithm $\mathsc{Up}\colon \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right)\times \bits^* \rightarrow \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right) \cup \{\bot\}$ takes a representation and an update string, and returns an updated representation or the distinguished symbol~$\bot$. 
%
The deterministic \emph{test} algorithm $\mathsc{Test} \colon \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right) \times U \rightarrow \bits$ takes a representation~$M$ and an element~$x \in U$ and returns a bit.
%
We assume that all $\mathsc{Rep},\mathsc{Qry},\mathsc{Up},\mathsc{Test}$ all have blackbox access to the functions $h_1,h_2,\ldots,h_k \in \mathcal{H}$, which we denote by writing~$\mathcal{H}$ as a superscript.

@@

\tsnote{Below is an alternative way to formalize things like counting Bloom filters, count-min sketch, scaling Bloom filters, etc.  Neither is solid, yet.}

\heading{Multiset-oriented hash-based filters. }
Here we extend the basic Bloom filter syntax to allow for variations on the traditional Bloom filter, e.g. counting Bloom filters.  \tsnote{We should find out what are the important variations! I only know of a few.  None of these require randomness, so I'll continue to formalize everything as deterministic.}\tsnote{“Scaling Bloom filters”, appear to allow the parameter~$m$, below, to vary. I've altered the syntax to allow this.}\tsnote{This syntax might naturally support count-min sketches, too.}

@@

An $(k,[m_1,m_2])$-filter with operations is a tuple  $B=(\mathsc{H},\mathsc{Rep},\mathsc{Qry}, \mathcal{F})$.  
The set $\mathcal{F}$ is the finite collection of allowable operations.  All operations are of the form 
$f: \mathbb{M}_{\mathcal{U}} \times \mathbb{M}_{\mathcal{U}} \rightarrow \mathbb{M}_{\mathcal{U}} \cup \{\bot\}$.  
%
The deterministic representation algorithm $\mathsc{Rep}\colon \mathbb{M}_\mathcal{U} \rightarrow \bigcup_{m=m_1}^{m_2}\Sigma^m$ takes a multiset~$S$, and outputs representation~$M$ of length~$m_1 \leq m m_2$.   
%
The randomized hash-sampling algorithm~$\mathsc{H}$ is as before.
%
The determinisitic query algorithm $\mathsc{Qry}$ \tsnote{Not sure how to define this!  See my comment, below...}

@@

\tsnote{There are more direct ways to formalize counting Bloom filters.  But this less direct way will allow us to capture other kinds ``advanced'' Bloom filters proposed in the literature or (more importantly) used in practice. For example, $f_{\mathrm{setify}}(S,T)=\{(x,1) \,|\, x \in \mathcal{U} \mbox{s.t. } \mu_S(x)>0, \mu_T(x)>0\}$.  On the other hand, we may end up deciding it is overkill... }

>>>>

\heading{Discusssion.}
This syntax captures classical Bloom filters by setting~$\Sigma=\bits$ and defining $\mathsc{Rep},\mathsc{Qry}$ in the obvious way.  Setting $\Sigma=\bits^{\leq L}$ allows one to store a bitstring of length at most~$L$ at each location. \tsnote{Would let us capture a construction from NY, among other things.}  We note that even though this would allow for the storage of counters (say) at each filter location, our restriction that~$\mathsc{Qry}$ returns a bit does not capture functionality of Bloom filters that employ counters, e.g., returning a counter value. (One could, however, have $\mathsc{Qry}$ compute an arbitrary, fixed predicate.  For example, ``has~$x$ been seen at least 5 times?''.)  Nonetheless, this restriction allows us to state a simple correctness condition, and covers a broad range of hash-based filters.

@@

\heading{Mutuable hash-based filters. }
Here we extend the basic Bloom filter syntax to allow for variations on the traditional Bloom filter, e.g. counting Bloom filters, spectral Bloom filters, count-min sketches.  \tsnote{We should find out what are the important variations! I only know of a few.  None of them require randomness for representation or querying, so I'll continue to formalize everything as deterministic.}

Fix a nonempty sets $U,\Sigma$ and integers $k,m_1,m_2,>0$ with $m_1 \leq m_2$.  Fix a symbol $\bot \not\in U$.  An $(k,[m_1,m_2])$-filter (over universe~$U$) is a tuple  $B=(\mathsc{H},\mathsc{Rep},\mathsc{Qry},\mathsc{Up}, \mathsc{Test})$.   
%
The randomized \emph{hash-sampling} algorithm~$\mathsc{H}$ samples a size~$k$ subset~$\mathcal{H} \subset \mathrm{Func}(U,\{0,1,\ldots,m_2-1\})$.  We write $\mathcal{H} \getsr \mathsc{H}$ for this operation.
%
The deterministic representation algorithm $\mathsc{Rep}\colon [U]^n \rightarrow \left(\bigcup_{m=m_1}^{m_2}\Sigma^m\right)$ takes a set~$S$ of~$n$ elements as input, and outputs representation~$M$ of length~$m_1 \leq m \leq m_2$.
%\mbox{s.t. } \mu_S(x)>0, \mu_T(x)>0\}$.  On the other hand, we may end up deciding it is overkill... }
%
The determinisitc query algorithm $\mathsc{Qry}\colon \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right)\times U \rightarrow \bits^*$ takes a representation $M$ and an element $x \in U$ and returns a bitstring.  
%
The deterministic \emph{update} algorithm $\mathsc{Up}\colon \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right)\times \bits^* \rightarrow \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right) \cup \{\bot\}$ takes a representation and an update string, and returns an updated representation or the distinguished symbol~$\bot$. 
%
The deterministic \emph{test} algorithm $\mathsc{Test} \colon \left(\bigcup_{m=m_1}^{m_2}\Sigma^m \right) \times U \rightarrow \bits$ takes a representation~$M$ and an element~$x \in U$ and returns a bit.
%
We assume that all $\mathsc{Rep},\mathsc{Qry},\mathsc{Up},\mathsc{Test}$ all have blackbox access to the functions $h_1,h_2,\ldots,h_k \in \mathcal{H}$, which we denote by writing~$\mathcal{H}$ as a superscript.

@@

\heading{Multiset-oriented hash-based filters. }
\tsnote{Below is an alternative way to formalize mutable hash-based filters.  This way directly address the inputs as multisets, instead of starting with a set and then updating the represenation a step at a time.  Not sure which is cleaner and more easily applied to real problems, yet.}

@@

An $(k,[m_1,m_2])$-filter with operations is a tuple  $B=(\mathsc{H},\mathsc{Rep},\mathsc{Qry}, \mathcal{F})$.  
The set $\mathcal{F}$ is the finite collection of allowable operations.  All operations are of the form 
$f: \mathbb{M}_{\mathcal{U}} \times \mathbb{M}_{\mathcal{U}} \rightarrow \mathbb{M}_{\mathcal{U}} \cup \{\bot\}$.  
%
The deterministic representation algorithm $\mathsc{Rep}\colon \mathbb{M}_\mathcal{U} \rightarrow \bigcup_{m=m_1}^{m_2}\Sigma^m$ takes a multiset~$S$, and outputs representation~$M$ of length~$m_1 \leq m \leq m_2$.   
%
The randomized hash-sampling algorithm~$\mathsc{H}$ is as before.
%
The determinisitic query algorithm $\mathsc{Qry}$... \tsnote{Not sure how to define this!  See my comment, below...}

@@

\tsnote{There are more direct ways to formalize counting Bloom filters, like the syntax above.  But this less direct way will allow us to capture other kinds ``advanced'' Bloom filters proposed in the literature or (more importantly) used in practice. For example, $f_{\mathrm{setify}}(S,T)=\{(x,1) \,|\, x \in \mathcal{U} \mbox{s.t. } \mu_S(x)>0, \mu_T(x)>0\}$.  On the other hand, we may end up deciding it is overkill... }
